---
title: "The Effect of Severe Weather on Public Health and the Economy"
output: html_document
---

This report investigates the effect storms and other severe weather events can cause both on public health and the economy of the United States. These severe events can result in fatalities, injuries, and property damage and finding key contributers can assist in the mitigation of such outcomes. My analysis has two parts. We see which extreme weather event has the greatest cost on, first, public health -- both fatalities and injuries and, second, on the U.S. economy.

## Data Processing
Just so paperwork is in order, these packages are required to perform the analysis.
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```

The data for this study is sourced from the National Climatic Data Center (NCDC), downloaded using the link found in the source code below (accessed 29/02/2015). The NCDC receives this data from the National Weather Service, who collate information from county, state and federal emergency management  officials, local law enforcement officials, skywarn spotters, etc. Documentation about how some of the variables are constructed/defined can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) (accessed 29/02/2015). The events in the database start in the year 1950 and end in November 2011.
```{r, echo=TRUE, message=FALSE, warning=FALSE, cache=TRUE}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", temp, mode = "wb")
StormData <- read.csv(temp)
unlink(temp)
```

## Results
In this section we investigate the data processed in the previous section. 

### Severe weather and public health
Here we determine the main contributing severe weather events that are harmful to public health. We assume that public health is affected by both fatalities and injuries and so both of these outcomes are added together and put in a new column as the variable "HARMED". We sum together those harmed with respect to each weather event and then list them in descending order.
```{r, echo=TRUE}
StormDataHarm <- select(StormData, EVTYPE, FATALITIES, INJURIES)
StormDataHarm <- mutate(StormDataHarm, HARMED = FATALITIES + INJURIES)

aggHarm <- aggregate(StormDataHarm$HARMED, list(Event = StormDataHarm$EVTYPE), sum, na.rm = TRUE)
aggHarm <- arrange(aggHarm, desc(x))
```

In Figure 1 we see the top five weather events that cause the most harm to public health in the United States from 1950 to 2011. It is clear that tornados are the biggest contributor, with 96,979 cases of death or injury, followed by excessive heat with `r aggHarm[2,2]` cases.
```{r, echo=TRUE}
ggplot(data = aggHarm[1:5,], aes(Event, x)) + geom_bar(stat="identity") + ylab("Number of cases")
```

### Severe weather and the economy
Here we investigate how severe weather events affect the United States economy. The columns selected are organized as follows: "EVTYPE" is the event type, "PROPDMG" and "CROPDMG" refer to damage of property and crops respectively. These are the coefficients to "PROPDMGEXP" and "CROPDMGEXP" which take on the values "K" or "k" for $1,000, "M" or "m" for $1,000,000 and "B" or "b" for $1,000,000,000. The expense for each extreme weather event is added up and sorted in descending order.
```{r, echo=TRUE}
StormDataExpense <- select(StormData, EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
StormDataExpense <- mutate(StormDataExpense, EXPENSE = 
                             {ifelse(PROPDMGEXP == "K" | PROPDMGEXP == "k", PROPDMG * 1000, 
                              ifelse(PROPDMGEXP == "M" | PROPDMGEXP == "m", PROPDMG * 1000000,
                              ifelse(PROPDMGEXP == "B" | PROPDMGEXP == "b", PROPDMG * 1000000000,
                                NA)))})

aggExpense <- aggregate(StormDataExpense$EXPENSE, list(Event = StormDataExpense$EVTYPE), sum, na.rm = TRUE)
aggExpense <- arrange(aggExpense, desc(x))
```

In Figure 2 we see the top 5 most expensive weather events in the United States from 1950 to 2011. It shows that Floods ("FLOO") have the most devastating effect, costing 145 billion dollars. This is followed by hurricanes/typhoons ("HURR") at 69 billion dollars.
```{r, echo=TRUE}
ggplot(data = aggExpense[1:5,], aes(Event, x)) + geom_bar(stat="identity") + scale_x_discrete(labels =  abbreviate)+ ylab("Cost in dollars")
```

